Improving disaggregation models of malaria incidence by ensembling non-linear models of prevalence

### 2019-12-13 tim-lucas-model-ensembles.tex
Add api refs.

The polygon incidence data (i.e. malaria incidence aggregated to administrative units) were collected from government reports and standardised using methods defined in \cite{cibulskis2011worldwide}.
This standardisation step accounts for missed cases due to lack of treatment seeking, missing case reports, and cases that sought medical attention outside the public health systems \citep{battle2016treatment}.
For reports where cases were not reported at the species level, national estimates of the ratio between \emph{P. falciparum} and \emph{Plasmodium vivax} cases were used to calculate \emph{P. falciparum} only cases. 
To minimise temporal effects we selected one year of surveillance data for each country. 
We used annual surveillance data from 2013 for Madagascar (110 districts), 2015 for Colombia (952 municipalities), 2013 for Indonesia (244 regencies and cities) and 2009 for Senegal (34 departments).
These years were selected as they had the most complete data in each case.

@@

For each country specific dataset and for the global dataset we fitted 5 stage 1 models via \emph{caret} \citep{caret}: elastic net \citep{enet}, Random Forest \citep{wright2015ranger}, projection pursuit regression \citep{friedman1981projection}, neural networks \citep{nnet} and boosted regression trees (gradient boosted models, subsequently GBM) \citep{gbm}.
These models were fitted to both the full malaria prevalence dataset and to the regional subsets of the data.
Our response variable was prevalence and we weighted the data by sample size (i.e.\thinspace the number of people tested for malaria in each survey).
We used the raw environmental covariates described above as covariates in these machine learning models.
This process can therefore be seen as creating non-linear transformations of the raw covariates that are hopefully better correlated with malaria incidence than the raw environmental covariates are.
For each model we ran five-fold cross-validation to select hyperparameters using random search for Random Forest and boosted regression trees and grid search for the other models. 
Root mean square error (RMSE) was used to select the best performing model.
Predictions from these models were then made across each country respectively.
These predictions were empirical logit transformed so that they were on the linear predictor scale of the disaggregation regression model.
An empirical logit was used rather than a standard logit as there were many predictions of exactly zero.
These predicted surfaces were subsequently used as covariates in the stage 2 dissaggregation regression models (Figure~\ref{f:schem}).
The experiments that follow in this paper assess the improved performance of disaggregation regression models when using the predictions from these machine learning models as covariates as compared to a baseline disaggregation regression model using only the raw environmental covariates.

@@

\centering
\begin{tabular}{llrrrr}
Country &  Cross-validation & Enviro &  Enviro + ML\textsubscript{l} &  ML\textsubscript{l} \\
\hline 
 Colombia & Random &  0.59 & \textbf{0.61}& 0.59  \\
 Colombia &  Spatial &  0.12 &  0.25&  \textbf{0.33} \\
 Indonesia &  Random &  0.52 &  \textbf{0.59}&  0.48  \\
 Indonesia &  Spatial &  0.45 &  \textbf{0.51}&  0.44  \\
 Madagascar & Random &   \textbf{0.70}& 0.69 &  0.68  \\
 Madagascar &  Spatial &  0.22 & 0.18&  \textbf{0.55} \\
 Senegal &  Random &  \textbf{0.58} & 0.57&  0.51  \\
 Senegal &  Spatial &  0.63 &  0.63&  0.51  \\
\end{tabular}
\caption{Pearson correlations between observed and predicted values. }
\label{t:results}

@@

\end{tabular}
\caption{Pearson correlations between observed and predicted values. }
\label{t:results2}

>>>>

The polygon incidence data (i.e. malaria incidence aggregated to administrative units) were collected from government reports \citep{mdg, sen, col, idn} and standardised using methods defined in \cite{cibulskis2011worldwide}.
This standardisation step accounts for missed cases due to lack of treatment seeking, missing case reports, and cases that sought medical attention outside the public health systems \citep{battle2016treatment}.
For reports where cases were not reported at the species level, national estimates of the ratio between \emph{P. falciparum} and \emph{Plasmodium vivax} cases were used to calculate \emph{P. falciparum} only cases. 
To minimise temporal effects we selected one year of surveillance data for each country. 
We used annual surveillance data from 2013 for Madagascar (110 districts), 2015 for Colombia (952 municipalities), 2013 for Indonesia (244 regencies and cities) and 2009 for Senegal (34 departments).
These years were selected as they had the most complete data in each case.

@@

For each country specific dataset and for the global dataset we fitted 5 stage 1 models via \emph{caret} \citep{caret}: elastic net \citep{enet}, Random Forest \citep{wright2015ranger}, projection pursuit regression \citep{friedman1981projection}, neural networks \citep{nnet} and boosted regression trees (gradient boosted models, subsequently GBM) \citep{gbm}.
These models were fitted to both the full malaria prevalence dataset and to the regional subsets of the data.
Our response variable was prevalence and we weighted the data by sample size (i.e.\thinspace the number of people tested for malaria in each survey).
We used the raw environmental covariates described above as covariates in these machine learning models.
This process can therefore be seen as creating non-linear transformations of the raw covariates that are hopefully better correlated with malaria incidence than the raw environmental covariates are.
For each model we ran five-fold cross-validation to select hyperparameters using random search for Random Forest and boosted regression trees and grid search for the other models. 
Root mean square error (RMSE) was used to select the best performing model.
Predictions from these models were then made across each country respectively.
These predictions were empirical logit transformed so that they were on the linear predictor scale of the disaggregation regression model.
An empirical logit was used rather than a standard logit as there were many predictions of exactly zero.
These predicted surfaces were subsequently used as covariates in the stage 2 dissaggregation regression models (Figure~\ref{f:schem}).
The experiments that follow in this paper assess the improved performance of disaggregation regression models when using the predictions from these machine learning models as covariates as compared to a baseline disaggregation regression model using only the raw environmental covariates.
See the supplementary material for plots of the grid search hyperparameter performance, out-of-sample scatter plots and plots of the predicted surfaces.

@@

\centering
\begin{tabular}{llrrrr}
Country &  Cross-validation & Enviro &  Enviro + ML\textsubscript{l} &  ML\textsubscript{l} \\
\hline 
 Colombia & Random &  0.59 & \textbf{0.61}& 0.59  \\
 Colombia &  Spatial &  0.12 &  0.25&  \textbf{0.33} \\
 Indonesia &  Random &  0.52 &  \textbf{0.59}&  0.48  \\
 Indonesia &  Spatial &  0.45 &  \textbf{0.51}&  0.44  \\
 Madagascar & Random &   \textbf{0.70}& 0.69 &  0.68  \\
 Madagascar &  Spatial &  0.22 & 0.18&  \textbf{0.55} \\
 Senegal &  Random &  \textbf{0.58} & 0.57&  0.51  \\
 Senegal &  Spatial &  0.63 &  0.63&  0.51  \\
\end{tabular}
\caption{Pearson correlations between observed and predicted values for experiment \emph{i}. }
\label{t:results}

@@

\end{tabular}
\caption{Pearson correlations between observed and predicted values for experiment \emph{ii}. }
\label{t:results2}
